---
title: "practical_machine_learning"
author: "SAIFUDDIN BIN MOHD"
date: "January 29, 2016"
output: html_document
---


###Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, we use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways.


###Data

The training data for this project are available here: [https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv]

The test data are available here: [https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv]


###Goal

The goal is to predict the manner in which they did the exercise. This is the ???classe??? variable in the training set. We use our prediction model to predict 20 different test cases in this project.


###Preliminary Work

An overall pseudo-random number generator seed was set at 1234 for all code. In order to reproduce the results below, the same seed should be used.
Packages were downloaded and installed is caret, randomForest, rpart and rpart.plot to be use in this project.

This report outcome variable is classe and factor variable with 5 levels. For this dataset, ???participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in 5 different fashions:

- Class A = really according to the specification 
- Class B = elbowed to the front
- Class C = lifting the dumbbell only halfway 
- Class D = dropping the dumbbell only halfway 
- Class E = throwing the hips to the front 

Class A corresponds to the specified execution of the exercise, while the other 4 classes correspond to common mistakes."
Prediction evaluations will be based on maximizing the accuracy and minimizing the out-of-sample error. All other available variables after cleaning will be used for prediction.
Two models will be tested using Decision Tree and Random Forest algorithms. The model with the highest accuracy will be selected.


####Expected sample error

The expected out-of-sample error will correspond to the quantity: 1-accuracy in the cross-validation data. Accuracy is the proportion of correct classified observation over the total sample in the subTesting data set. Expected accuracy is the expected accuracy in the out-of-sample data set (i.e. original testing data set). Thus, the expected value of the out-of-sample error will correspond to the expected number of missclassified observations/total observations in the Test data set, which is the quantity: 1-accuracy found from the cross-validation data set.


####Cross-validation

It will be performed by subsampling our training data set randomly without replacement into 2 subsamples: subTraining data (75% of the original Training data set) and subTesting data (25%). Our models will be fitted on the subTraining data set, and tested on the subTesting data. Once the most accurate model is choosen, it will be tested on the original Testing data set.


####Reasons for my choices

Our outcome variable ???classe??? is an unordered factor variable. Thus, we can choose our error type as 1-accuracy. Large sample size with N = 19622 in the Training data set. So it were divided into subTraining and combine with subTesting to allow cross-validation. Columns or fields with all missing values will be discarded as well as features that are irrelevant. All other features will be kept as relevant variables.
Random Forest and Decision Tree algorithms are known for their ability of detecting the features that are important for classification.


###Code and Results

Installing packages, loading libraries, and setting the seed for reproduceability:

```{r, echo=FALSE}
library(caret)
library(randomForest) #Random forest for classification and regression
library(rpart) # Regressive Partitioning and Regression trees
library(rpart.plot) # Decision Tree plot
```
setting the overall seed for reproduceability
```{r}
set.seed(1234)
```

Loading the data sets into R and irrelevant variables will be deleted.

Missing values are coded as string "#DIV/0!" or "" or "NA" - these will be changed to NA.

We notice that both data sets contain columns are irrelevant to produce this report, these will be deleted.  

Loading the training data set and replacing all missing with "NA"
```{r}
trainingData <- read.csv("pml-training.csv", na.strings=c("NA","#DIV/0!", ""))
```

Loading the testing data set and replacing all missing with "NA"
```{r}
testingData <- read.csv('pml-testing.csv', na.strings=c("NA","#DIV/0!", ""))
```

Check dimensions for number of variables and number of observations
```{r}
dim(trainingData)
dim(testingData)
```

Delete columns with all missing values
```{r}
trainingData<-trainingData[,colSums(is.na(trainingData)) == 0]
testingData <-testingData[,colSums(is.na(testingData)) == 0]
```

Some variables are irrelevant to our current project: user_name, raw_timestamp_part_1, raw_timestamp_part_,2 cvtd_timestamp, new_window, and  num_window (columns 1 to 7). We can delete these variables.
```{r}
trainingData <-trainingData[,-c(1:7)]
testingData <-testingData[,-c(1:7)]
```

and have a look at our new datasets:
```{r}
dim(trainingData)
dim(testingData)
```
Heading content for trimmed datasets (without columns 1-7) can be shown in Appendix 

Partitioning the training data set to allow cross-validation

The training data set contains 53 variables and 19622 obs.
The testing data set contains 53 variables and 20 obs.
In order to perform cross-validation, the training data set is partionned into 2 sets: subTraining (75%) and subTesting (25%).
This will be performed using random subsampling without replacement.

```{r}
subSamples <- createDataPartition(y=trainingData$classe, p=0.75, list=F)
subTraining <- trainingData[subSamples, ] 
subTesting <- trainingData[-subSamples, ]
dim(subTraining)
dim(subTesting)
```
Heading content for subsample datasets can be shown in Appendix 

The variable ???classe??? have 5 levels: A, B, C, D and E. To show the frequency difference, a plot of classe levels vs frequency can be plot. 
```{r}
plot(subTraining$classe, col=rainbow(5), main="Levels of the variable classe within the subTraining datasets", xlab="Classe Levels", ylab="Frequency")
```

The graph show "Level A is the most frequent with more than 4000 occurrences while "Level D is the least frequent which few extra from 2000 occurrences.

First prediction model: Decision Tree
```{r}
model1 <- rpart(classe ~ ., data=subTraining, method="class")
```
Predicting:
```{r}
predict1 <- predict(model1, subTesting, type="class")
```
Plot of the Decision Tree
```{r}
rpart.plot(model1, main="Classification Tree", extra=102, under=TRUE, faclen=0)
```

Test results on our subTesting data set:
```{r}
confusionMatrix(predict1, subTesting$classe)
```

Second prediction model: Random Forest
```{r}
model2 <- randomForest(classe ~. , data=subTraining, method="class")
```

Predicting:
```{r}
predict2 <- predict(model2, subTesting, type = "class")
```
Test results on subTesting data set:
```{r}
confusionMatrix(predict2, subTesting$classe)
```


###Conclusion

As a conclusion, Random Forest algorithm performed better than Decision Trees which is
accuracy for Random Forest model was 0.995 compared to Decission Tree model was 0.739. The expected outside sample error is estimated at 0.005, or 0.5%. The expected outside sample error is calculated as 1 - accuracy for predictions made against the cross-validation set (0.995). With an accuracy above 99% on our cross-validation data, we can expect that very few, or none, of the test samples will be notclassified.

Output submission
predict outcome levels on the original Testing data set using Random Forest algorithm
```{r}
predictFinal <- predict(model2, testingData)
predictFinal
```


###Appendix
Heading content for trimmed datasets (without columns 1-7) :
```{r}
head(trainingData)
head(testingData)
```

Heading content for subsample datasets :
```{r}
head(subTraining)
head(subTesting)
```

```{r}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("case_no_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(predictFinal)
```